---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```
## Introduction

The `ProtPipe` package provides downstream proteomics workflows based on `SummarizedExperiment`. This vignette uses a packaged example object so the analysis starts from a ready-to-use dataset rather than reconstructing the object from raw files.

## Setup and Data Loading

First, we load the `ProtPipe` package.
```{r setup}
suppressPackageStartupMessages(library(ProtPipe2))
library(SummarizedExperiment)
library(ggplot2) # For plot customization
```

Load the packaged example object with `data()`. This object was created from `EXAMPLES/basic_example_data/iPSC.csv` without supplying external sample metadata, so the only sample annotation present initially is `differentiation_day`.

```{r load-data}
data("protpipe_example_se")
se <- protpipe_example_se

se
colData(se)
```

## Initial Quality Control (QC)

Before normalization and analysis, we should assess the quality of our data.

### Protein Counts and Intensity Distributions

We can check the number of proteins identified in each sample and visualize the intensity distributions with boxplots.

```{r qc-plots}
# Get the number of identified proteins per sample
ProtPipe2::get_pg_counts(se)

# Plot the counts for each sample
ProtPipe2::plot_pg_counts(se)

# Plot the intensity distributions for each sample
ProtPipe2::plot_pg_intensities(se)
```
These plots help us identify any samples that behave as strong outliers.

### Sample Correlation

Next, we can assess the reproducibility between replicates by plotting a correlation heatmap. Samples from the same condition should generally cluster together.

```{r correlation-heatmap}
ProtPipe2::plot_correlation_heatmap(se)
```

## Data Pre-processing

`ProtPipe` includes functions for common pre-processing steps like normalization and imputation.

### Normalization
Here, we apply median normalization to align the intensity distributions across all samples.
```{r normalization}
se_normalized <- ProtPipe2::median_normalize(se)

# We can re-plot the intensities to see the effect of normalization
ProtPipe2::plot_pg_intensities(se_normalized)
```

### Imputation
Missing values must be handled before many downstream analyses. We will use the "down-shifted normal" (Perseus-like) imputation method. As this is a stochastic method, we set a seed for reproducibility.

```{r imputation}
set.seed(123) # Set a seed for reproducible imputation
se_imputed <- ProtPipe2::impute_left_dist(se_normalized)

# The object should no longer have missing values
any(is.na(assay(se_imputed)))

# Create a preprocessing report 
report <- generate_preprocessing_report(se_imputed)
```


## Downstream Analysis and Visualization

With a clean, complete dataset, we can now explore the relationships between samples and proteins.

### Principal Component Analysis (PCA)

PCA is a powerful tool for visualizing the primary sources of variation in the data and assessing sample clustering.

```{r pca-plot}
# The plot_pca function is a convenient wrapper that calculates and plots the results
ProtPipe2::plot_PCs(se_imputed, condition = "differentiation_day") +
  labs(title = "PCA by time point")
```

### UMAP

Lets plot a UMAP.

```{r umap-plot}
# The plot_pca function is a convenient wrapper that calculates and plots the results
ProtPipe2::plot_umap(se_imputed, condition = "differentiation_day", neighbors = 6) +
  labs(title = "UMAP by time point")
```

### Differential Expression

To illustrate a simple two-group comparison, we compare Day 28 and Day 0
directly with limma.

```{r volcano-plot}
de <- ProtPipe2::do_limma_binary(
  se_imputed,
  condition = "differentiation_day",
  control_group = "Day0",
  treatment_group = "Day28"
)

ProtPipe2::plot_volcano(
  de,
  label_col = "PG.Genes"
)
```

### Pathway Analysis

We can perform a simple Gene Ontology enrichment analysis on the differential
expression results.

```{r pathway-analysis}
pathways <- ProtPipe2::enrich_pathways(
  de,
  gene_col = "PG.Genes",
  source = "go",
  run_gsea = FALSE,
  run_kegg = FALSE
)

pathways$plots$ora_up_dotplot
```

### Protein Expression Heatmap

Finally, we can visualize the expression patterns of the proteins across our samples using a heatmap. The data is automatically Z-scored by row to highlight relative expression changes.

```{r final-heatmap}
# Plot a heatmap of all proteins with row and column clustering
ProtPipe2::plot_proteomics_heatmap(
  se_imputed,
  protmeta_col = "PG.Genes",
  condition = "differentiation_day",
  cluster_rows = TRUE,
  cluster_cols = TRUE
)

top_genes <- unique(stats::na.omit(de$PG.Genes))[1:4]

ProtPipe2::plot_proteomics_heatmap(
  se_imputed,
  protmeta_col = "PG.Genes",
  condition = "differentiation_day",
  genes = top_genes,
  cluster_rows = TRUE,
  cluster_cols = TRUE
)
```

## Conclusion

This vignette demonstrated a complete ProtPipe workflow starting from a
packaged `SummarizedExperiment`. The same pattern extends to user-supplied
objects constructed with `create_se()`.
