---
title: "getting-started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{getting-started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## Introduction

The `ProtPipe` package provides a streamlined, start-to-finish workflow for common proteomics data analysis tasks. It is built around a central `ProtData` object, which stores quantitative data, protein metadata, and sample metadata in a single, cohesive unit. This vignette will walk you through a typical analysis, from data loading to quality control and visualization.

## Setup and Data Loading

First, we load the `ProtPipe` package.
```{r setup}
library(ProtPipe)
library(SummarizedExperiment)
library(ggplot2) # For plot customization
```

Our analysis begins with two primary data frames:

1.  A `data.frame` containing protein metadata and quantitative intensity data for each sample.
2.  A `data.frame` containing the experimental design, mapping each sample to its condition.

```{r create-data}
# 1. Create a sample raw data frame
# 1. Create a more realistic sample raw data frame
# We will simulate 50 genes in total:
# - 40 genes that are stable (no change between conditions)
# - 5 genes that are up-regulated in the Treatment group
# - 5 genes that are down-regulated in the Treatment group
n_stable <- 400
n_up <- 50
n_down <- 50
n_total <- n_stable + n_up + n_down

# Set means for each group for log2-scale data
mean_stable <- 25
mean_up <- 51  # Higher expression
mean_down <- 12 # Lower expression

# Note: set.seed() makes the random data reproducible
set.seed(123) 

# Simulate data for each sample group
data <- data.frame(
  Ctrl_1 = rnorm(n_total, mean = mean_stable, sd = 2),
  Ctrl_2 = rnorm(n_total, mean = mean_stable, sd = 2.5),
  Ctrl_3 = rnorm(n_total, mean = mean_stable, sd = 2.2),
  Treat_1 = c(rnorm(n_stable, mean_stable, 2), rnorm(n_up, mean_up, 2), rnorm(n_down, mean_down, 2)),
  Treat_2 = c(rnorm(n_stable, mean_stable, 2.5), rnorm(n_up, mean_up, 2.5), rnorm(n_down, mean_down, 2.5)),
  Treat_3 = c(rnorm(n_stable, mean_stable, 2.2), rnorm(n_up, mean_up, 2.2), rnorm(n_down, mean_down, 2.2))
)

# Introduce some missing values (NAs) to make it more realistic
# Set ~5% of the intensity values to NA
num_cells <- ncol(data) * n_total
na_indices <- sample(num_cells, size = round(0.05 * num_cells))
data <- as.matrix(data)
data[na_indices] <- NA
data <- as.data.frame(data)

# Combine into a single data frame with protein metadata
raw_data <- data.frame(
  ProteinID = paste0("P", 1:n_total),
  Gene = paste0("GENE", 1:n_total),
  Status = c(rep("Stable", n_stable), rep("Upregulated", n_up), rep("Downregulated", n_down)),
  data
)

# 2. Create the corresponding condition data frame
cond_df <- data.frame(
   SampleID = c("Ctrl_1", "Ctrl_2", "Ctrl_3", "Treat_1", "Treat_2", "Treat_3"),
   group = c("Control", "Control", "Control", "Treatment", "Treatment", "Treatment"),
   batch = c("A", "B", "A", "B", "A", "B")
)

# 3. Use the constructor to create the ProtData object
pd_obj <- ProtPipe::create_se(data = raw_data, sample_metadata = cond_df)

# We can inspect the created object
print(pd_obj)
```

## Initial Quality Control (QC)

Before normalization and analysis, we should assess the quality of our data.

### Protein Counts and Intensity Distributions

We can check the number of proteins identified in each sample and visualize the intensity distributions with boxplots.

```{r qc-plots}
# Get the number of identified proteins per sample
ProtPipe::get_pg_counts(pd_obj)

# Plot the counts for each sample
ProtPipe::plot_pg_counts(pd_obj)

# Plot the intensity distributions for each sample
ProtPipe::plot_pg_intensities(pd_obj)
```
These plots help us identify any samples that behave as strong outliers.

### Sample Correlation

Next, we can assess the reproducibility between replicates by plotting a correlation heatmap. Samples from the same condition should generally cluster together.

```{r correlation-heatmap}
ProtPipe::plot_correlation_heatmap(pd_obj, order_by = "group", label_by = "group")
```

## Data Pre-processing

`ProtPipe` includes functions for common pre-processing steps like normalization and imputation.

### Normalization
Here, we apply median normalization to align the intensity distributions across all samples.
```{r normalization}
pd_obj_normalized <- ProtPipe::median_normalize(pd_obj)

# We can re-plot the intensities to see the effect of normalization
ProtPipe::plot_pg_intensities(pd_obj_normalized)
```

### Imputation
Missing values must be handled before many downstream analyses. We will use the "down-shifted normal" (Perseus-like) imputation method. As this is a stochastic method, we set a seed for reproducibility.

```{r imputation}
set.seed(123) # Set a seed for reproducible imputation
pd_obj_imputed <- ProtPipe::impute_left_dist(pd_obj_normalized)

# The object should no longer have missing values
any(is.na(assay(pd_obj_imputed)))

# Create a preprocessing report 
report <- generate_preprocessing_report(pd_obj_imputed)
```


## Downstream Analysis and Visualization

With a clean, complete dataset, we can now explore the relationships between samples and proteins.

### Hierarchial Clustering (PCA)

```{r hc-plot}
# The plot_pca function is a convenient wrapper that calculates and plots the results
ProtPipe::plot_hierarchical_cluster(pd_obj_imputed) +
  labs(title = "HC of Samples by Group")
```
### Principal Component Analysis (PCA)

PCA is a powerful tool for visualizing the primary sources of variation in the data and assessing sample clustering.

```{r pca-plot}
# The plot_pca function is a convenient wrapper that calculates and plots the results
ProtPipe::plot_PCs(pd_obj_imputed, condition = "group") +
  labs(title = "PCA of Samples by Group")
```

### UMAP

Lets plot a UMAP.

```{r umap-plot}
# The plot_pca function is a convenient wrapper that calculates and plots the results
ProtPipe::plot_umap(pd_obj_imputed, condition = "group", neighbors = 4) +
  labs(title = "UMAP of Samples by Group")
```
### Differential Expression

Lets figure out which proteins are altered by the treatment.

```{r volcano plot}
DE <- ProtPipe::do_limma_binary(pd_obj_imputed, condition = "group", control_group = "Control", treatment_group = "Treatment")

# Or, filter to a specific subset of genes of interest
ProtPipe::plot_volcano(
  DE, 
  label_col = "Gene"
)
```
### Protein Barchart

Lets take a closer look at some of the differentially expressed proteins

```{r barchart}
# Plot a heatmap of all proteins, with columns ordered by clustering
ProtPipe::compare_protein(pd_obj_imputed, prot = "GENE465", prot_meta_col = "Gene", condition = "group")
```


### Protein Expression Heatmap

Finally, we can visualize the expression patterns of the proteins across our samples using a heatmap. The data is automatically Z-scored by row to highlight relative expression changes.

```{r final-heatmap}
# Plot a heatmap of all proteins, with columns ordered by clustering
ProtPipe::plot_proteomics_heatmap(pd_obj_imputed, protmeta_col = "Gene")

# Or, filter to a specific subset of genes of interest
ProtPipe::plot_proteomics_heatmap(
  pd_obj_imputed, 
  protmeta_col = "Gene", 
  genes = c("GENE1", "GENE2", "GENE3", "GENE4")
)
```

## Conclusion

This vignette demonstrated a complete analysis workflow using the `ProtPipe` package. By combining data and metadata into a `ProtData` object, we can apply a series of QC, pre-processing, and visualization steps in a clear and reproducible pipeline.
